---
title: "02 Data wrangling"
---

# 1) Setup  

```{r Setup, message=F}
# Loading packages (every time you open R)
library(readxl) # to read excel files
library(janitor) # to clean data
library(dplyr) # wrangling
library(tidyr) # wrangling
library(readr) # to export csv

```


```{r Reading data}
soyk = read_xlsx(path ="../data/01_soybeans.xlsx") # using the relative path (../) to import or export data

# Checking first rows
soyk

# View
View(soyk)
```

This study was conducted at four different locations in one year to assess the effect of potassium (K) fertilizer rates on soybean trifoliate K concentration at growth stage R4.  

Each study was a randomized complete block design (RCBD) with K rate as the only treatment factor with levels 0, 50, 100, and 150 lbs K2O/ac.

```{r Glimpse}
glimpse(soyk) # similar to structure (str)
```

```{r Summary}
summary(soyk)
```

# 2) Data wrangling  
```{r soyk_1 - clean_names()}
soyk_1 = clean_names(soyk)

soyk_1
```


```{r soyk_2 - filter()}
names(soyk_1)

soyk_2 = filter(soyk_1, 
                k_rate_lbac != 0
                
                )

soyk_2
```

```{r soyk_3 - mutate()}
# Transforming K rate from lbs/ac to kg/ha
soyk_3  = mutate(soyk_1,
                 k_rate_kgha = k_rate_lbac*0.454/0.405,
                 k_rate_kgha = round(k_rate_kgha, 0)
                 )

soyk_3
```

```{r soyk_4 - case_when()}
soyk_4 = mutate(soyk_3,
                rep = case_when(
                  plot > 100 & plot < 200 ~ 1,
                  plot > 200 & plot < 300 ~ 2,
                  plot > 300 & plot < 400 ~ 3,
                  plot > 400 & plot < 500 ~ 4
                  
                )
                ) %>%
  mutate(plot_2 = as.character(plot)) %>%
  mutate(plot_2 = stringr::str_extract(plot_2, ""))

soyk_4

```

```{r soyk_5 - pivot_longer()}
names(soyk_4)

soyk_5 = pivot_longer(soyk_4,
                      names_to = "location.k",
                      values_to = "k_pct",
                      cols = contains("_k_pct") # matching the pattern of the cols to be changed to long format
                      
                      ) # convert wide data format to long format

soyk_5

View(soyk_5)
```

```{r soyk_6 - separate()}
soyk_6 = separate(soyk_5,
                  location.k,
                  into = c("location", "k", "pct")
                  )

soyk_6

View(soyk_6)
```

```{r soyk_7 - select()}
names(soyk_6)

soyk_7 = dplyr::select(soyk_6,
                location, rep, k_rate_kgha, k_pct
                
                )

soyk_7
```

```{r soyk_8 - arrange()}
soyk_8 = arrange(soyk_7, location)

soyk_8
```

```{r soyk_9 and 10 - group_by and summarise}
# group_by
soyk_9 = group_by(soyk_8, location)

soyk_9

class(soyk_8)

class(soyk_9)

# summarise
soyk_10 = summarise(soyk_9,
                    mean_pct = mean(k_pct)
                    
                    )

soyk_10
```

```{r Piping it all together}
# Use data wrangling functions in a pipe workflow to generate the average k_pct for each K rate (in kg/ha) within each location (basically, averaging over reps)

soyk_w = soyk %>%
  clean_names() %>%
  mutate(
    k_rate_kgha = round(k_rate_lbac * 0.454 / 0.405, 0),
    rep = case_when(
      plot > 100 & plot < 200 ~ 1,
      plot > 200 & plot < 300 ~ 2,
      plot > 300 & plot < 400 ~ 3,
      plot > 400 & plot < 500 ~ 4
    )
  ) %>%
  pivot_longer(
    cols = contains("k_pct"),
    names_to = "location_k",
    values_to = "k_pct"
  ) %>%
  separate(
    location_k,
    into = c("location", "k", "pct"),
    sep = "_"
  ) %>%
  dplyr::select(location, rep, k_rate_kgha, k_pct) %>%
  arrange(location, k_rate_kgha, rep) # %>%
  # group_by(location, k_rate_kgha) %>%
  # summarise(
  #   mean_k_pct = mean(k_pct, na.rm = TRUE),
  #   .groups = "drop"
  # )


soyk_w


```

```{r Checking if same as without pipe}
# Same result?
identical(soyk_8, soyk_w)

```

```{r Saving to file for further analysis}


```

# 3) Assignment   
Starting from soyk, use data wrangling functions in a pipe workflow to generate the average k_pct for each K rate (in kg/ha) within each location (basically, averaging over reps). You can use the chunk below to complete this quiz.

```{r assignment}
soyk

soyk_k = soyk %>%
  clean_names() %>%
  mutate(
    k_rate_kgha = round(k_rate_lbac * 0.454 / 0.405, 0),
    rep = case_when(
      plot > 100 & plot < 200 ~ 1,
      plot > 200 & plot < 300 ~ 2,
      plot > 300 & plot < 400 ~ 3,
      plot > 400 & plot < 500 ~ 4
    )
  ) %>%
  pivot_longer(
    cols = contains("k_pct"),
    names_to = "location_k",
    values_to = "k_pct"
  ) %>%
  separate(
    location_k,
    into = c("location", "k", "pct"),
    sep = "_"
  ) %>%
  dplyr::select(location, rep, k_rate_kgha, k_pct) %>%
  arrange(location, k_rate_kgha, rep) %>%
  group_by(location, k_rate_kgha) %>%
  summarise(
    mean_k_pct = mean(k_pct, na.rm = TRUE),
    .groups = "drop"
  )

View(soyk_k)
```


Tip: the final data frame will have 16 rows.

Submit your result (take a screenshot that includes the code and the chunk output) on eLC under Assignment #2 by Monday Jan 26th 11:59 pm.



